---
title: "R packaging"
author: "Emy Guilbault & Arthur Rodrigues"
date: "2023-10-12"
output: 
  xaringan::moon_reader:
    css:  ["default", "metropolis-fonts"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, echo=FALSE}
.blackbox {
  padding: 1em;
  background: lightgray;
  color: grey;
  border: 2px solid darkorange;
  border-radius: 10px;
}
.center {
  text-align: center;
}
/* Two-column layout */
.left-column {
  color: #777;
  width: 20%;
  height: 92%;
  float: left;
}
.left-column h2:last-of-type, .left-column h3:last-child {
  color: #000;
}
.right-column {
  width: 75%;
  float: right;
  padding-top: 1em;
}
.pull-left {
  float: left;
  width: 47%;
}
.pull-right {
  float: right;
  width: 47%;
}
.pull-right + * {
  clear: both;
}

/* Metropilis style*/
.remark-slide-content {
  background-color: #FAFAFA;
  border-top: 80px solid #23373B;
  font-size: 20px;
  font-weight: 300;
  line-height: 1.5;
  padding: 1em 2em 1em 2em
}

.title-slide .inverse .remark-slide-content {
  background-color: #FAFAFA;
}

.inverse {
  background-color: #23373B;
  text-shadow: none;
}

/* Removes colored bar from top of the slide resulting in a clear slide */
.clear{
  border-top: 0px solid #FAFAFA;
}

h1 {
  font-weight: normal;
  margin-top: -95px;
  margin-left: -00px;
  color: #FAFAFA;
}

h2, h3, h4 {
  padding-top: -15px;
  padding-bottom: 00px;
  color: #1A292C;
  text-shadow: none;
  font-weight: 400;
  text-align: left;
  margin-left: 00px;
  margin-bottom: -10px;
}

.remark-slide-content h1 {
  font-size: 45px;
}

.remark-slide-content h2 {
  font-size: 35px;
}

.remark-slide-content h3 {
  font-size: 30px;
}

.left-column h2, .left-column h3, .left-column h4 {
  color: #777;
}

.left-column h2:last-of-type, .left-column h3:last-child {
  color: #1A292C;
}

.title-slide {
  background-color: #FAFAFA;
  border-top: 80px solid #FAFAFA;
}

.title-slide h1  {
  color: #1A292C;
  font-size: 40px;
  text-shadow: none;
  font-weight: 400;
  text-align: left;
  margin-left: 15px;
  padding-top: 80px;
}
.title-slide h2  {
  margin-top: -25px;
  padding-bottom: -20px;
  color: #1A292C;
  text-shadow: none;
  font-weight: 300;
  font-size: 35px;
  text-align: left;
  margin-left: 15px;
}
.title-slide h3  {
  color: #1A292C;
  text-shadow: none;
  font-weight: 300;
  font-size: 25px;
  text-align: left;
  margin-left: 15px;
  margin-bottom: -30px;
}

hr, .title-slide h2::after, .mline h1::after {
  content: '';
  display: block;
  border: none;
  background-color: #EB811B;
  color: #EB811B;
  height: 1px;
}

hr, .mline h1::after {
  margin: 1em 15px 0 15px;
}

.title-slide h2::after {
  margin: 10px 15px 35px 0;
}

.mline h1::after {
  margin: 10px 15px 0 15px;
}

.remark-slide-number {
  font-size: 13pt;
  color: #272822;
  opacity: 1;
}
.inverse .remark-slide-number {
  font-size: 13pt;
  color: #FAFAFA;
  opacity: 1;
}

/* turns off slide numbers for title page: https://github.com/gnab/remark/issues/298 */
.title-slide .remark-slide-number {
  display: none;
}

.remark-inline-code {
  /* background: #F5F5F5; /* lighter */
  background: #e7e8e2; /* darker */
  border-radius: 3px;
  padding: 4px;
}

.code10 .remark-code {
  font-size: 10%;
}

.code20 .remark-code {
  font-size: 20%;
}

.code30 .remark-code {
  font-size: 30%;
}

.code40 .remark-code {
  font-size: 40%;
}

.code50 .remark-code {
  font-size: 50%;
}

.code60 .remark-code {
  font-size: 60%;
}

.code70 .remark-code {
  font-size: 70%;
}

.code80 .remark-code {
  font-size: 80%;
}

.code90 .remark-code {
  font-size: 90%;
}

.code100 .remark-code {
  font-size: 100%;
}

.font10 {
  font-size: 10%;
}

.font20 {
  font-size: 20%;
}

.font30 {
  font-size: 30%;
}

.font40 {
  font-size: 40%;
}

.font50 {
  font-size: 50%;
}

.font60 {
  font-size: 60%;
}

.font70 {
  font-size: 70%;
}

.font80 {
  font-size: 80%;
}

.font90 {
  font-size: 90%;
}

.font100 {
  font-size: 100%;
}

.font110 {
  font-size: 110%;
}

.font120 {
  font-size: 120%;
}

.font130 {
  font-size: 130%;
}

.font140 {
  font-size: 140%;
}

.font150 {
  font-size: 150%;
}

.font160 {
  font-size: 160%;
}
.font170 {
  font-size: 170%;
}
.font180 {
  font-size: 180%;
}
.font190 {
  font-size: 190%;
}
.font200 {
  font-size: 200%;
}

/* Various left and right split columns*/

.pull-left-70 { float: left; width: 66.5%; }
.pull-right-30 { float: right; width: 28.5%; }
.pull-right-30 + * { clear: both; }

.pull-left-60 { float: left; width: 57%; }
.pull-right-40 { float: right; width: 38%; }
.pull-right-40 + * { clear: both; }

.pull-left-50 { float: left; width: 47.5%; }
.pull-right-50 { float: right; width: 47.5%; }
.pull-right-50 + * { clear: both; }

.pull-left-40 { float: left; width: 38%; }
.pull-right-60 { float: right; width: 57%; }
.pull-right-60 + * { clear: both; }

.pull-left-30 { float: left; width: 28.5%; }
.pull-right-70 { float: right; width: 66.5%; }
.pull-right-70 + * { clear: both; }

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

```


#  Presentation: Motivation and use (ARTHUR)
$${\color{red}Why? \space what \space do \space you \space need? }$$
      &rarr; If you need to tailor some existing functions, if you are going to use them many times. If you want to save time and have a way to standardize your coding with existing one.
  
$${\color{orange}Who \space is \space it \space for?}$$
      &rarr; Are the changes or the created function of any use for other coders or anyone working in similar field?
  
$${\color{gold}Where?}$$ 
      &rarr; do you need a place to share code and make it available. github is an option. CRAN make it more accessible.
  
$${\color{green}How?}$$ 
      &rarr; We are going to dig into the steps of creating a package!

---
# Presentation: Motivation and use (ARTHUR)
Here we go beyond the agreed knowledge that everyone knows broadly what is a package and how to install new packages: using `install.packages("x")` and then `library(X)`.
  
  
---
# Presentation: Motivation and use (ARTHUR)

One of the main principles in good coding practice is 'do not repeat yourself'. There are some options to achieve these goal. One of them is to create a loop, when your task is only to repeat a procedure with different input. Other option is to create a custom function, which allow you to modify predefined places of your code using arguments. 
  
When we are creating a analysis or cleaning data it could be useful to write a function and let it in a separated file. Then, you only need to add a `source()` in your script to load your custom function in the R environment and you are able use it across your script. Using this approach will make your code more readable and tidy, because instead of several lines of code to do a step of your analysis, you now have only few lines of code using your custom function. Moreover, you can use it several times in your script without need to copy and paste your code and change few characters. 
  
Therefore, for a project you are working, you can create as much functions you need, and load it your scripts using the `source()` function. However, if you have created a set of functions that could be useful in other projects or for other people, it could be nice to create an R package. Which could also let your coding more susceptible to errors. 
  
---
# Pros and cons to create a package
Since you have some custom functions, you might wonder if it is a good idea to create a package. 
  
Pros:   
- Easy access to functions that you've created (it is easier to add a `library()` call in your script than a `source()`)  
- Your functions have help pages
- Your functions could be used by others
- Users could give feedback to your function 
  
Cons:  
- Could be time consuming to create and maintain a package
- Users could give feedback to your functions (bugs)




---
# Package structure:  (Emy)
---

What compose the R package?
  
* DESCRIPTION file: This is the metadata for the package. It contains:
  + Package name
  + Title and description
  + Authors
  + Dependencies (depends, imports, suggests)
  + Licensing
  + Version number
  + Bug report location, and so on
  
* R/ directory for R files that contain your functions
  
* NAMESPACE file which specifies the function that compose your package and what imports functions from othe rpacages are needed.
  
These are created using:  `usethis::create_package()`
  
* Other optional information such as data etc..
---

The Description file is a simple file format called DCF, the Debian control format. It is created using `usethis::create_package("mypackage")`.
For example, we will set up the different fields:

.blackbox[
.font60[Package: Gradients

Title: Creates a gradient image and modify it. *only one line*

Version: 0.0.0.9000 *consider carefully the version*

Authors@R: 
    person("First", "Last", , "first.last@example.com", role = c("aut", "cre"),
           comment = c(ORCID = "YOUR-ORCID-ID"))

Description: Gradients creates an customized image that display a gradient. Other 
functionalities include modifying the radient direction or stepness as well as transposing directly the gradient image. *What the package does (one paragraph)*

License: `use_mit_license()`, `use_gpl3_license()` or friends to pick a
    license

Encoding: UTF-8

Roxygen: list(markdown = TRUE)

RoxygenNote: 7.2.3]
]

---


### R code :   
Let's consider few function for raster and image modification. First, our new functions and worflow will depend on other packages that we need to load:
```{r, echo=TRUE, warning=FALSE, message = FALSE}
library(spatstat)
library(lattice)
```

Next, we want to create an image that display a gradient along the area of interest. Thus set the coordinates of our gridded area (resolution) and define a function which will evaluate the values $v1$ across our space. The last step is to create the image using `as.im()`.
---

.pull-left-60[
```{r, echo=T, eval=F}
set.seed(113)
XY = expand.grid(seq(0, 100, 1), 
                 seq(0, 100, 1))
X = XY[,1]
Y = XY[,2]

v1 = (X - 30)^2 + (Y + 80)^2 - 0.5*X*Y
v1 = -1*scale(v1)

raster.im = as.im(data.frame(x = X,
                             y = Y, 
                             z = v1))
plot(raster.im)
```
]

.pull-right-40[
```{r, echo=F, out.width="80%"}
set.seed(113)
XY = expand.grid(seq(0, 100, 1), seq(0, 100, 1))
X = XY[,1]
Y = XY[,2]

v1 = (X - 30)^2 + (Y + 80)^2 - 0.5*X*Y
v1 = -1*scale(v1)

raster.im = as.im(data.frame(x = X, y = Y, z = v1))
plot(raster.im)
```
]

---
# Gradientfunction

---
# Transposefunction


---
# wigglefunction


---
# Creation and Maintenance (Arthur)
.pull-left-90[
Main packages needed for package development:
* devtools: Creation - building of the package
* usethis: help with documentation, testing and broadcast
* roxygen2: help with the documentation set up
* pkgdown: for building a site for your package
* testthat: testing

--

- One bullet

- Another bullet

--

--

! You also need a system builder ! Make sure Rtools is installed which is done differently depending on your system (windows, MacOS, Linux).
]


---
# Step by step
Whole R package development workflow:

* `available::available("pkgname")` # check if package name is available (if planning to publish publicly)
* `usethis::create_package("pkgname")` # create the package
* `usethis::use_git()` # set up version control
* `usethis::use_github()` # optional
* `usethis::use_r("myfile")` # creating new R file in the R/ directory
* `usethis::use_data_raw()` # if adding data
* `devtools::load_all()` # try it out in the console
* `usethis::use_package("import-pkgname")` # add package to import (or depends or suggests)


---
# Step by step
Documentations:

* `usethis::use_package_doc()` # add package documentation
* `usethis::use_pipe()` # if you want to add %>% from `magrittr`
* `usethis::use_vignette("vignette-name")` # add vignette
* `usethis::use_test()` # make test file for active R file




---
# Step by step
Write some test

* `devtools::test_active_file()` # test active file
* `devtools::test()` # test whole package
* `devtools::build()` # build vignettes
* `devtools::install()` # to install package
* `devtools::check()` # to build and check a package 
* `usethis::use_readme_rmd()` # to add a README Rmd file
* `styler::style_pkg()` # optional (commit first, then review changes carefully)
* `usethis::use_pkgdown_github_pages()` # for setting up pkgdown website on github
* `usethis::use_pkgdown()` if not using github pages

---
# Bugs and recommendations

---
# Summary of resources

https://r-pkgs.org/description.html

https://rpubs.com/emitanaka/rladies-lagos-paris-2023

https://github.com/EmyGlblt/Rpackaging


---

class: center, middle

.Large[Thanks!]

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).